Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-03-28 2024pipit/figures/core-structure.tex
entailment updates
copy and tighten

%!TEX root = ../Main.tex
\begin{figure}
  \includegraphics[width=\textwidth]{figures/core-structure.pdf}
\caption{Architecture of Pipit. The gray boxes and solid arrows are defined in this paper. The white boxes and dashed arrows are trusted components. The labels next to the arrows denote properties of the translation. The refinement label ($\subseteq$) indicates a verified abstraction relation; the negated entailment ($\not\vdash$) indicates that the entailment relation --- contracts and checked properties proven on abstract system apply to the checked semantics --- is not yet verified; the equivalence ($=$) denotes a verified equivalence relation.}
\label{f:core:structure}
\end{figure}

>>>>

%!TEX root = ../Main.tex
\begin{figure}
  \includegraphics[width=\textwidth]{figures/core-structure-1920x450.pdf}
\caption{Architecture of Pipit. The gray boxes and solid arrows are defined in this paper. The white boxes and dashed arrows are trusted components. The labels denote verified properties of the translation: abstraction~($\subseteq$), entailment of proof obligations~($\vdash$), and equivalence~($=$).
}
\label{f:core:structure}
\end{figure}

### 2024-03-28 2024pipit/section/01-Introduction.tex
copy and tighten

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
To verify programs, Pipit translates its expression language to a transition system for k-inductive proofs, which is verified to be an abstraction of the original semantics.
To execute programs, Pipit can generate executable code, which is total and semantics-preserving.

@@

\begin{itemize}
  \item we motivate the need to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions rather than relations; composing functions results in simpler transition systems (\autoref{s:transition});
  \item we identify the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:core:causality}, \autoref{s:transition:proof});
  \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
  \item finally, we evaluate Pipit by implementing the high-level logic of a time-triggered Controller Area Network (CAN) bus driver, which we have partially verified (\autoref{s:evaluation}).
\end{itemize}

>>>>

To be confident that our proofs hold on the real system, we need a single shared semantics for the compiler and the prover.
In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
To verify programs, Pipit translates its expression language to a transition system for k-inductive proofs, which is verified to be an abstraction of the original semantics.
To execute programs, Pipit can generate executable code, which is total and semantics-preserving.

@@

\begin{itemize}
  \item we motivate the need to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit; after checking deferred properties, programs are \emph{blessed}, which marks their properties as valid (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions rather than relations; composing functions results in simpler transition systems (\autoref{s:transition});
  \item we identify the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:core:causality}, \autoref{s:transition:proof});
  \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
  \item finally, we evaluate Pipit by implementing the high-level logic of a Time-Triggered Controller Area Network (TTCAN) bus driver and verifying an abstract model of a key component (\autoref{s:evaluation}).
\end{itemize}

### 2024-03-28 2024pipit/section/20-Motivation.tex
copy and tighten

At a high level, the network schedule is described by a \emph{system matrix} which consists of rows of \emph{basic cycles}.
Each basic cycle consists of a sequence of actions to be performed at particular time-marks.
Actions in the schedule may not be relevant to all nodes, so the node's \emph{node matrix} contains only the relevant actions.
The node matrix is represented in-memory by a constant \emph{triggers array} containing the relevant triggers, sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
Reference messages start a new basic cycle; a subset of nodes, designated as leaders, send reference messages to synchronise the network.
Watch alerts are generally placed after an expected reference message and are used to signal an error if no reference message is received.

@@

We first implement a streaming function \emph{count_when} to maintain the index into the triggers array; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
At each time-mark, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.

@@

% The trigger-fetch logic requires such a combination of automatic and manual proofs.
We can now implement the trigger-fetch logic, which keeps track of the current trigger.
The trigger-fetch logic uses the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.

@@

With the explicitly instantiated lemma, we can prove the streaming invariant by straightforward induction on the transition system.
To help compose this function with the rest of the system, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
The contract we state is that if the environment ensures that the time progresses as expected --- that is, we are called once per time-mark --- then we guarantee that we never encounter a late trigger.

@@

The key distinction between these streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one microsecond after the time at the previous tick of computation.
Expressing such a rely in an imperative setting would require extra encoding, as imperative specification languages do not generally have an innate notion of discrete time steps.

>>>>

At a high level, the network schedule is described by a \emph{system matrix} which consists of rows of \emph{basic cycles}.
Each basic cycle consists of a sequence of actions to be performed at specific time-marks.
Actions in the schedule may not be relevant to all nodes; the node's \emph{node matrix} contains only the relevant actions.
The node matrix is represented by a constant \emph{triggers array} containing triggers sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
Reference messages start a new basic cycle; a subset of nodes, designated as leaders, send reference messages to synchronise the network.
Watch alerts are generally placed after an expected reference message to signal an error if no reference message is received.

@@

We implement a streaming function \emph{count_when} to maintain the index into the triggers array; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
At each step, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the counter as-is.

@@

% The trigger-fetch logic requires such a combination of automatic and manual proofs.
We can now implement the trigger-fetch logic, which keeps track of the current trigger.
We use the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
We simplify our presentation here and only consider a constant cycle: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.

@@

With the explicitly instantiated lemma, we can prove the streaming invariant by straightforward induction on the transition system.
To help compose this function with the rest of the system, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
The contract we state is that if we are called once per time-mark then we guarantee that we never encounter a late trigger.
% The contract we state is that if the environment ensures that the time progresses as expected --- that is, we are called once per time-mark --- then we guarantee that we never encounter a late trigger.

@@

The key distinction between our streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one time-mark after the time at the previous tick of computation.
Expressing such a rely in an imperative setting would require extra encoding, as imperative specification languages do not generally have an innate notion of discrete time steps.

### 2024-03-28 2024pipit/section/30-Core.tex
ok
copy and tighten

% \pagebreak
\section{Core language}
\label{s:core}

@@

\autoref{f:core:structure} shows the high-level architecture of Pipit.
On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
The translation from the surface syntax to the core language is trusted.
There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
The translation to abstract systems also generates proof obligations for the checked properties, contract definitions and contract instantiations in the core program; these proof obligations are verified to correspond to the original properties.
% One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
The translation to executable transition systems is proven to be semantics-preserving, as is the subsequent translation to \lowstar{}.
The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.

@@

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero upwards, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
% We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
% Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a record or tuple.

@@

For example, the core expression $\sumrec$ computes the sum of values from a stream of integers $i$ by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
$$
\sumcontract
$$
To be considered a valid program, we would have to prove that the contract definition itself holds, similar to our earlier contract in \autoref{s:motivation:contract}.
The unknown property status here allows us to defer the caller's proof that the input stream is always positive until we know more about the actual call-site.

@@

To prove this instantiation valid, we need to show that the input $(\xfby{1}{\fibvar})$ is positive.
To do this, we perform induction on the input streams, allowing us to assume that the rely was previously true.
We then use the contract to show that, if the rely was true at the previous step, then the guarantee was true at the previous step -- which implies that the previous result is positive.
We will describe this proof in more detail in \autoref{s:transition:ind}.

>>>>

% \pagebreak
\section{Pipit core language}
\label{s:core}

@@

\autoref{f:core:structure} shows the high-level architecture of Pipit.
On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
The translation from the surface syntax to the core language is trusted.
There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
The translation to abstract systems also generates proof obligations, which are verified to correspond to the proof obligations on the original program.
% One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
The translation to executable transition systems is proven to be semantics-preserving, as is the subsequent translation to \lowstar{}.
The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.

@@

Recursive streams are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero up, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
% We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
% Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a record or tuple.

@@

For example, the core expression $\sumrec$ computes the sum of values from a stream of integers $i$ by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
$$
\sumcontract
$$
To be considered a valid program, we must prove that the contract definition itself holds, as with our earlier contract (\autoref{s:motivation:contract}).
The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.

@@

To prove this instantiation valid, we need to show that the input $(\xfby{1}{\fibvar})$ is positive.
To do this, we perform induction on the input streams, allowing us to assume that the rely was previously true.
We then use the contract to show that, if the rely was true at the previous step, then the guarantee was true at the previous step -- which implies that the previous result is positive.
We describe the mechanisms of this proof in \autoref{s:transition}.
